---
title: "Marketing Effectiveness Study - Fruitazia"
author: "Njabulo Hlabangana"
date: "2022-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The aim of this exercise is to assess the effectiveness of marketing expenditure on the sales of an organization called Fruitazia using standard statistical techniques like linear regression and analysis of variance. The case study was part of the Marketing Analytics Course offered by Columbia University on Edx. I took part in the auditing track of the course.



## Loading the requisite packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggpubr) #anova visualization
```

## Loading the data

The data presents results of a market test for fruit juice with 2 advertising levels × 3 price levels, which gives 6 experimental conditions, with 5 supermarkets per condition.

```{r}
fruitazia <- readxl::read_excel("C:/Users/hlaba/Desktop/Marketing/fruitazia.xlsx")
fruitazia$Price <- as.factor(fruitazia$Price) # converting price to a factor 
fruitazia$AdSpend <- as.factor(fruitazia$AdSpend) # converting AdSpend into a factor

fruitazia
```
In the data, each out of 30 rows encodes experiment description per supermarket, including set price and advertising expenditure as well as the resulting sales. Sales are in thousands of cases. Price is in dollars. Ad spend is in million dollars. The rest of the variables are dummies encoding distinct price, ad spend levels, and their interactions.

## Data summary

The code below shows the mean sales by for each price level and associated ad spending.

```{r, message= FALSE, warning= FALSE}
library(knitr)
library(kableExtra)
kable(aggregate(Sales ~ Price + AdSpend, data=fruitazia, FUN=mean))
```

The results of the test are depicted graphically below. The general observation is that sales tend to go down as prices go increases. This is basically micro-economics 101. Also, for any given price level, more advertising expenditure tends to generate higher sales.This is shown by the 2 million dollar AdSpend curve being above the 1 million dollar one at all levels of the price factor. 

```{r, message= FALSE, warning=FALSE}
ggline(fruitazia, x = "Price", y = "Sales", color = "AdSpend",
       add = c("mean_se", "dotplot"))
```

## Statistical Analysis

The descriptive analysis above points to a significant effect of both price and advertisement spending on sales. The question, however, is whether or not the effect is statistically significant and this section seeks to answer that question.

**One-way Analysis of Variance** (ANOVA) will be used first. The *lm()* function is used to ascertain the significance of the all the factor levels which are encoded as dummy variables. All the factor levels are statistically significant at the 0.1% level of significance.

```{r}
model <- lm(Sales ~ Price + AdSpend, data=fruitazia)
summary(model)
```

The analysis above shows the effect of the individual levels of the Price and AdSpend factor levels but does not quite indicate the overall effects of the Price and AdSpend factors. Calling the *ANOVA()* function helps with that as used below.

```{r}
anova(model)
```
The anova analysis confirms the overall statistical significance of both Price and AdSpend factors. They are both significant at the 0.1% level.

## ANOVA with Interaction

A quick look at the graphical depiction of the effect of AdSpend shown in the graph above shows that the effect of AdSpend gets lower as the price increases. This is shown by the shrinking of the space between the two lines as the price level increases and can be an indicator of the possible interaction between Price and AdSpend. If AdSpend affected Sales the same way across all Price levels, the gap between the two lines in the graph would be more or less constant - the two lines would be parallel.

It therefore necessary to conduct a statistical test to check for the possible statistical significance of this interaction as done below.

```{r}
model_interaction <- lm(Sales ~ Price * AdSpend, data=fruitazia)
summary(model_interaction)
```
 
 The effect of interaction effect is negative and statistically significant at the 5% level only for the highest price of $5.9 per unit. This corroborates the graphical analysis finding that Adspend is less effective at higher higher prices. 